---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
``` {r}
activity=read.csv("activity.csv")
# change column 2 from factor to date
activity$date <- as.Date(activity$date, "%Y-%m-%d")
library(dplyr)
```
The activity data contains 17568 counts for the number of steps taken in a 5 mintue interval during 61 days in October and November 2012:
``` {r}
head(activity)
```

## Part 1: Ignoring Missing Values
## What is the total number of steps taken per day?
``` {r}
# remove missing values
# Note: Using na.rm=TRUE instead would result in an incorrect sum of 0 for days with all NA values
activity1 <- activity[!is.na(activity$steps),]
# calculate total number of steps per day
by_day <- summarize(group_by(activity1,date), sum(steps))
names(by_day) <- c("date", "steps")
# create historgram
hist(by_day$steps, col = "red", main = "Total Steps Taken per Day", xlab = "Number of Steps")
```

## What is mean and median total number of steps taken per day?
``` {r}
# calculate mean number of steps per day
mean_steps <- sprintf("%g",mean(by_day$steps))
# calcuate the median number of steps per day
median_steps <- sprintf("%g",median(by_day$steps))
```
The mean number of steps is `r mean_steps`

The median number of steps is `r median_steps`

## What is the average daily activity pattern?
``` {r}
# calcuate the average number of steps per 5 minute interval
by_interval <- summarize(group_by(activity1,interval), mean(steps))
names(by_interval) <- c("interval", "steps")
# create plot
plot(by_interval$interval,by_interval$steps, type="l",col = "blue", xlab = "5 Mintue Interval", ylab = "Average Number of Steps", main = "Average Number of Steps Across All Days for 5 Minute Intervals")
```

## Which 5-minute interval contains the maximum number of steps?
``` {r}
# find the interval with the maximum number of steps
max_index <- which.max(by_interval$steps)
max_interval <- by_interval$interval[max_index]
```
The 5-minute interval containing the maximum average number of steps is `r max_interval`

## Part 2: Imputing Missing Values
## What are the total number of missing values in the dataset?
``` {r}
# calculate the total number of NAs
missing <- is.na(activity$steps)
num_nas <- sum(missing)
```
The total number of missing values in the dataset is `r num_nas`

##Devise a strategy for filling in all of the missing values in the dataset:
Choose one of the folowing:

- mean for that day
- median for that day
- mean for the 5-minute interval

First determine which days and 5-minute intervals have missing values
``` {r}
# group NAs by date and count number of intervals in each group
by_missing <- summarize(group_by(activity[missing,],date),count=n())
by_missing
```
There are 8 days that have missing values for all of the 5-minute intervals,
so a mean or median for that day cannot be computed.  
Therfore use the mean for the 5-minute interval.

## Use 5-minute interval means to fill in the missing values:
``` {r}
activity2 <- activity
# replace missing values using the 5-minute interval means
for (i in 1:nrow(activity2)) {
    if (is.na(activity2$steps[i])) {
        activity2$steps[i] <- round(by_interval$steps[by_interval$interval==activity2$interval[i]])
    }
}
head(activity2)
```

## What is the total number of steps taken per day?
``` {r}
# calculate total number of steps per day using imputed values
by_day <- summarize(group_by(activity2,date), sum(steps))
names(by_day) <- c("date", "steps")
# create historgram
hist(by_day$steps, col = "red", main = "Total Steps Taken per Day", xlab = "Number of Steps")
```

## What is mean and median total number of steps taken per day?
``` {r}
# calculate mean number of steps per day
mean_steps2 <- sprintf("%g",mean(by_day$steps))
# calcuate the median number of steps per day
median_steps2 <- sprintf("%g",median(by_day$steps))
```
The mean number of steps is `r mean_steps2` (compared to `r mean_steps` from Part 1)

The median number of steps is `r median_steps2` (compared to `r median_steps` from Part 1)

These values are only slightly different from the values in Part 1 (where missing values were ignored), and therefore imputing values using the mean for the 5-minute interval has negligible impact.

## Are there differences in activity patterns between weekdays and weekends?
``` {r}
# create a new factor variable in the dataset indicating whether a given date is a weekday or weekend day
activity2 <- mutate(activity2,day=factor(grepl("^S",weekdays(date)),labels=c("weekday","weekend")))
# calcuate the number of steps per 5 minute interval averaged across all weekdays or weekend days
by_interval <- summarize(group_by(activity2,interval,day), mean(steps))
names(by_interval) <- c("interval", "day", "steps")
# make a panel plot containing the weekend and weekday data
library(lattice)
xyplot(steps ~ interval | day, data = by_interval, type = "l", layout = c(1, 2), xlab="5 Minute Interval", ylab="Average Number of Steps")
```

The above plots show some difference in activity patterns between weekdays and weekends. The weekday plot shows a siginificant activity spike in the morning, followed by much lower levels throughout the remaining waking hours, whereas the activity level during the waking hours on the weekend is more uniform.